---
title: "Analysis of severe weather events on public health and economic consequences across United States"
author: "Maverix13"
date: "September 21, 2015"
output: pdf_document
keep_md: yes
---

## Synopsis
This report explores  the NOAA Storm Database and answer some basic questions about severe weather events. 
The analysis covers:

* Across United States, events that are most harmful to public health.
* Across United States, events that have greatest economic consequences.

Results, covered later in the document, shows that 50% of the fatalities are caused by Extreme Temperature and Convection related weather events while combined with Flood and Wind accounts for about 80% of fatalities. Similarly, Convection and Extreme Temperatures cause about 60% of injuries, which combined with combined with Flood and Wind causes over 80% injuries. Economic impact of weather related activities is mostly caused by Flood and Cyclone resulting in over $250 billion. Property damage accounts for over 90% of the economic impact. 

## Data Processing

Data is downloaded, if required. Code checks if data already exists. Data is then loaded into R. 
Note that a zip file is directly read using read.csv. There is no need to unzip it as read.csv will handle it.
```{r message=FALSE}
#install.packages("devtools")
#devtools::install_github("renkun-ken/formattable")
library(knitr)
library(dplyr)
library(ggplot2)
library(formattable)
library(reshape2)
require(gridExtra)
```
```{r cache=TRUE}
if(!file.exists("data")){
  dir.create("data")
}
if(!file.exists("data/stormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/stormData.csv.bz2", method = "curl")
  
}

stormData <- read.csv("data/stormData.csv.bz2")
dim(stormData)
```
```{r}
head(stormData[,2:28])
```

### Preprocessing Data
As the histogram below shows, in the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. 
```{r cache=TRUE}
stormData <- mutate(stormData, YEAR = as.numeric(format(as.Date(BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y")))
g <- ggplot(stormData, aes(YEAR, fill = ..count.., removePanelGrid=TRUE,removePanelBorder=TRUE)) +
  geom_histogram(binwidth=5) + 
  ylab("Number of Events") +
  scale_fill_gradient("# Events", low = "black", high = "gray") +
  ggtitle("Histogram: Events By Year")
print(g)
```

Hence for this report, data after 1996 is considered for analysis.

```{r cache=TRUE}
stormData <- stormData[stormData$YEAR >= 1996,]
dim(stormData)
```

Following columns are of relevance for the current analysis

Column Name | Description
------------|------------
EVTYPE      | Type of storm event. Take note that similar storm events can be listed using different wording e.g. “coastal flood” and “coastal flooding.” Take note of this if you want to run a query grouping by event type. 
FATALITIES  | Number directly killed
INJURIES    | Number directly injured
PROPDMG		  |	Property damage in whole numbers and hundredths
PROPDMGEXP	|	A multiplier where Hundred (H), Thousand (K), Million (M), Billion (B)
CROPDMG		  |	Crop damage in whole numbers and hundredths
CROPDMGEXP	|	A multiplier where Hundred (H), Thousand (K), Million (M), Billion (B)

This above table is constructed based on information from http://ire.org/nicar/database-library/databases/storm-events/

Further reduction of stormData based on columns and row values

```{r cache=TRUE}
## Just keeping relevant columns
relevantColumns <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
stormData <- stormData[,relevantColumns]
## Keep rows having any data
stormData <- subset(stormData, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)


```

#### Mapping of Events to Group
Many events can be grouped together based on similarity. The categorization for this report is based on NOAA 2009 Annual Summaries (http://www.ncdc.noaa.gov/oa/climate/sd/annsum2009.pdf).

```{r}
stormData$EVGROUP <- "Others"
stormData[grepl("tornado|thunderstorm|hail|funnel|tstm|lig(.*?)ing", 
                stormData$EVTYPE, ignore.case = TRUE), "EVGROUP"] <- "Convection"
stormData[grepl("temperature|cold|cool|heat|hot|fire|dry|warm|wet|freez", 
                stormData$EVTYPE, ignore.case = TRUE), "EVGROUP"] <- "Extreme Temperature"
stormData[grepl("flood|rain|mud(.*?)slid", 
                stormData$EVTYPE, ignore.case = TRUE), "EVGROUP"] <- "Flood"
stormData[grepl("marine|sea|tide|coast|tsunami|current|surf|wave", 
                stormData$EVTYPE, ignore.case = TRUE), "EVGROUP"] <- "Marine"
stormData[grepl("tropic|cyclone|hurricane", 
                stormData$EVTYPE, ignore.case = TRUE), "EVGROUP"] <- "Cyclone"
stormData[grepl("avalanche|snow|blizzard|winter|wintry|ice", 
                stormData$EVTYPE, ignore.case = TRUE), "EVGROUP"] <- "Winter"
stormData[grepl("wind|dust", 
                stormData$EVTYPE, ignore.case = TRUE), "EVGROUP"] <- "Wind"

stormData$EVGROUP <- as.factor(stormData$EVGROUP)
```

### Analysis on public health
Analysis of fatalities and injuries caused by a given storm event group. 

```{r cache=TRUE}
pubHealthData <- stormData %>% 
  group_by(EVGROUP) %>% 
  summarize(fatalities = sum(FATALITIES), injuries = sum(INJURIES))
kable(pubHealthData)
```

### Analysis on Economic Impact
Analysis of property and crop damage caused by a given event group. 
```{r}
unique(stormData$PROPDMGEXP)
unique(stormData$CROPDMGEXP)
```
Unique values for both property and crop damage exponents is ['K', 'M', 'B']. These values are mulitpliers as mentioned above in column descriptions. 

```{r cache=TRUE}
economicData <- stormData %>% 
  mutate(PROPDMGVAL = PROPDMG * (
    ifelse(PROPDMGEXP == "K", 1000,
           ifelse(PROPDMGEXP == "M", 1000000,
                  ifelse(PROPDMGEXP == "B", 1000000000,
                         1  )))))  %>% 
  mutate(CROPDMGVAL = CROPDMG * (
    ifelse(CROPDMGEXP == "K", 1000,
           ifelse(CROPDMGEXP == "M", 1000000,
                  ifelse(CROPDMGEXP == "B", 1000000000,
                         1  ))))) %>%
  mutate(TOTALDMG = PROPDMGVAL + CROPDMGVAL) %>%
  group_by(EVGROUP) %>%
  summarize(propertyDamage = sum(PROPDMGVAL), cropDamage = sum(CROPDMGVAL), totalDamage = sum(TOTALDMG)) %>%
  arrange(desc(totalDamage))
economicData$EVGROUP <- factor(economicData$EVGROUP, levels=economicData$EVGROUP)
formattable(economicData, list(
  propertyDamage=currency,
  cropDamage=currency,
  totalDamage=currency
))
```

## Results

### Impact on public health
```{r}
fData <- pubHealthData %>%
  arrange(desc(fatalities)) %>%
  select(EVGROUP, fatalities) %>%
  mutate(
    cumulativeSum = cumsum(fatalities),
    cumPercent = cumulativeSum/sum(fatalities)
  )
formattable(fData, list(
  cumPercent = percent
))
```

Table above shows fatalities along with a cumulative percent. Extreme Temperature, Convection, Flood and Wind together cause over 7500 fatalities which is close to 80% of total fatalities. 
```{r}
iData <- pubHealthData %>%
  arrange(desc(injuries)) %>%
  select(EVGROUP, injuries) %>%
  mutate(
    cumulativeSum = cumsum(injuries),
    cumPercent = cumulativeSum/sum(injuries)
  )
formattable(iData, list(
  cumPercent = percent
))
```

As with fatalities, injuries are also mostly caused by Convection, Extreme Temperature, Flood and Wind resulting in over 50,000 injuries.
```{r}
fPlot <- ggplot(fData, aes(x=reorder(EVGROUP, desc(fatalities)))) +
  geom_bar(aes(y=fatalities), stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Severe Weather Category") +
  scale_y_continuous("Number of Fatalities") +
  ggtitle("Number of Fatalities\n by Weather Category\n from 1996 - 2011")
iPlot <- ggplot(iData, aes(x=reorder(EVGROUP, desc(injuries)))) +
  geom_bar(aes(y=injuries), stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Severe Weather Category") +
  scale_y_continuous("Number of Injuries") +
  ggtitle("Number of Injuries\n by Weather Category\n from 1996 - 2011")
grid.arrange(fPlot, iPlot, ncol = 2)
```

Based on the graphs above, we see that there are lot more injuries compared to fatalities but the top 2 weather event are Extreme Tmperature and Convection.

### Impact on Economy

```{r}
pData <- economicData %>%
  arrange(desc(propertyDamage)) %>%
  select(EVGROUP, propertyDamage) %>%
  mutate(
    cumulativeSum = cumsum(propertyDamage),
    cumPercent = cumulativeSum/sum(propertyDamage)
  )
formattable(pData, list(
  propertyDamage=currency,
  cumPercent = percent
))
```

Table above shows that over $250 billion in prpoerty damage is caused by Flood and Cyclone.

```{r}
cData <- economicData %>%
  arrange(desc(cropDamage)) %>%
  select(EVGROUP, cropDamage) %>%
  mutate(
    cumulativeSum = cumsum(cropDamage),
    cumPercent = cumulativeSum/sum(cropDamage)
  )
formattable(cData, list(
  cropDamage=currency,
  cumPercent = percent
))
```

Comparing crop damage to property damage shows that crop damage does not have as significant economic impact as property damage. Crop damage totaled close to $34 billion across all weather events. Hence the chart below takes total property damage into account.

```{r}

meltData <- melt(economicData[,c('EVGROUP','propertyDamage','cropDamage')], id.var="EVGROUP")
ePlot <- ggplot(meltData, aes(x = EVGROUP, y = value/1000000000, fill = variable)) + 
  geom_bar(stat = "identity") +
  xlab("Severe Weather Category") +
  scale_y_continuous("Damage $ (Billions of Dollars)") +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(ePlot) 

```

Majority of economic impact is attributed to property damage caused by Flood and Cyclone.

